On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-11 Manuscript/manuscript.tex
Additional citation

%3
In 2009, the International Association of Athletics Federations (Also referred to
as IAAF or World Athletics), which is the governing board for international running
competitions, commissioned a study into changing the reaction time barrier.
The paper analyzed sprinters reaction times and the biological processes 
that went into reacting to the starting gun and concluded 
that athletes can react in as low as 0.08 seconds.  One study by Pain looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. Another study commissioned
by IAAF concluded that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
\citep{komi2009iaaf} This could provide a more accurate representation of 
reaction time, and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.
A 2012 follow-up study to Komi's by Pilianidis found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  This contradicts
what was found in a 2021 study that concluded the year of competition has a
significant effect on reaction times \citep{zhang2021correlation}.  Zhang's
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, athletes reaction times got
faster. However, not  all studies have been in favor of a decrease in the
reaction time barrier. Brosnan in 2016 argues for increasing the reaction time 
barrier to 0.115 seconds, and cited evidence from European Championships and 
World Championships in his article \citep{brosnan2017effects}.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
 react in as little as 0.1 seconds \citep{lipps2011implications}.
It is worth noting that one study looking at data from various IAAF meets found
that the fastest reaction times for men were found between the ages of 26 and 29
\citep{tonnessen2013reaction}.  This is notable because Allen was 27 years old
at the 2022 Track and Field World Championship and in his physical peak.

>>>>

%3
In 2009, the International Association of Athletics Federations (Also referred to
as IAAF or World Athletics), which is the governing board for international running
competitions, commissioned a study into changing the reaction time barrier.
The paper analyzed sprinters reaction times and the biological processes 
that went into reacting to the starting gun and concluded 
that athletes can react in as low as 0.08 seconds.  One study by Pain looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. Another study commissioned
by IAAF concluded that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
\citep{komi2009iaaf} This could provide a more accurate representation of 
reaction time, and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.
A 2012 follow-up study to Komi's by Pilianidis found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  This contradicts
what was found in a 2021 study that concluded the year of competition has a
significant effect on reaction times \citep{zhang2021correlation}.  Zhang's
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, athletes reaction times got
faster. However, not  all studies have been in favor of a decrease in the
reaction time barrier. Brosnan in 2016 argues for increasing the reaction time 
barrier to 0.115 seconds, and cited evidence from European Championships and 
World Championships in his article \citep{brosnan2017effects}.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.01 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 Track and Field World Championship 
and in his physical peak.
